---
title: "Homework 2"
author: "Andrea Vasquez"
date: " 10/04/2026"
output: 
  github_document:
    toc: true
---

# Problem 1

### 1.1 Data import and cleaning + brief description 

```{r include=FALSE}
library(tidyverse)
```


```{r}

NYC_transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("NA", "", "."),
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character()))|> 
  janitor::clean_names()|> 
  select(line, station_name, station_latitude, station_longitude,
         starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))

NYC_transit_df

skimr::skim(NYC_transit_df)
```

The NYC transit dataset contains information on each subway entrance and exit. After cleaning, it has `r nrow(NYC_transit_df)`, rows and `r ncol(NYC_transit_df)` columns, of which 15 are character variables (line, station name, routes 1–11, vending, and entrance type), 2 are logical variables (entry and ADA compliance), and 2 are numeric variables (station latitude and longitude). To clean the datam, I imported the file specifying that blank cells be treated as missing and that routes 8-11 be read as character variables. These columns contain numbered trans and would otherwise be read as numeric, even though the values are rout name and not true numbers. I then cleaned the variables names, retained only the variables of interest, and converted the entry variable from character to logical. Given that route spread across 11 columns, we can conclude that this data is not tidy. To be tidy, there should be a single column for the variable route. 

### 1.2 Answering questions

#### Q1

```{r}
n_stations = 
  NYC_transit_df|>
  distinct(line, station_name)|>
  nrow()
```

There are `r n_stations` distinct stations. 

#### Q2

```{r}
n_ada_compliant =
  NYC_transit_df|>
  filter(ada == TRUE)|>
  distinct(line, station_name)|>
  nrow()
```

In total, `r n_ada_compliant` stations are compliant with ADA. 

#### Q3

```{r}
prop_entry_ven = 
  NYC_transit_df|>
  filter(vending == "NO")|>
  pull (entry)|>
  mean()
```

Among entrances/exits without vending, `r round(prop_entry_ven * 100, 2)`% allow entrance. 

### 1.3 Reformat data 

```{r echo=TRUE}

NYC_transit_tidy = 
  NYC_transit_df |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name")|> 
  drop_na(route_name)

NYC_transit_tidy
```

### 1.4 Questions about train A

```{r}

n_serving_A = 
  NYC_transit_tidy|>
  filter(route_name == "A")|>
  distinct(line, station_name)|>
  nrow()

n_A_compliant = 
  NYC_transit_tidy|>
  filter(route_name == "A", ada == TRUE)|>
  distinct(line, station_name)|>
  nrow()
```

There are `r n_serving_A` stations serving the A train, of which `r n_A_compliant` are compliant with ADA. 

# Problem 2

### 2.1 Data import and cleaning

```{r}

Mr_Wheel_df = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742",
    na = c("NA", "", "."))|> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.integer(year), 
    trash_wheel = "Mr_Trash") |> 
  relocate(trash_wheel)

Prof_Wheel_df = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    na = c("NA", "", ".", "dive"))|> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.integer(year), 
    trash_wheel = "Professor_Trash") |> 
  relocate(trash_wheel)

Gwynnda_df = 
  readxl::read_excel(
    "data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394",
    na = c("NA", "", "."))|> 
  janitor::clean_names() |> 
  drop_na(dumpster) |> 
  mutate(
    year = as.integer(year), 
    trash_wheel = "Gwynnda") |> 
  relocate(trash_wheel)

trash_wheel_df = 
  bind_rows(Mr_Wheel_df, Prof_Wheel_df, Gwynnda_df)

trash_wheel_df
```

### 2.2 Data set description

The combined dataset contains `r nrow(trash_wheel_df)` observations and `r ncol(trash_wheel_df)` variables. The data were collected from `r min(pull(trash_wheel_df, year))` to `r max(pull(trash_wheel_df, year))`, and each observation represents a dumpster filled by one of three trash wheels: Mr. Trash Wheel (`r nrow(Mr_Wheel_df)` dumpsters), Professor Trash Wheel (`r nrow(Prof_Wheel_df)` dumpsters), and Gwynnda (`r nrow(Gwynnda_df)` dumpsters). Some key variables include the trash wheel that collected the dumpster (`trash_wheel`), the dumpster number and the date of collection, the weight (in tons) and volume (in cubic yards) of trash collected, as well as the count of specific items, such as plastic bottles, polystyrene, cigarette butts, plastic bags, sports balls, and wrappers. Some variables are not recorded for every wheel. For example, sports balls are only available for Mr. Trash Wheel. 

It is important to note that given some quality issues present in the raw data, the combined data may not look exactly the same. For instance, the Gwynnda sheet skips the dumpster numbers 373 to 379 and dumpster 21 appears twice, resulting in a total number of dumpster (`r nrow(Gwynnda_df)`) different from the highest recorded number (`r max(pull(Gwynnda_df, dumpster))`). In the Professor Trash Wheel sheet, one entry had the text "dive" in place of a plastic bottle count and so was treated as a missing value.


### 2.3 Answer to Questions: 

```{r}
prof_total_weight = 
  trash_wheel_df |> 
  filter(trash_wheel == "Professor_Trash") |> 
  pull(weight_tons) |> 
  sum(na.rm = TRUE)

gwynnda_butts_june22 = 
  trash_wheel_df |> 
  filter(trash_wheel == "Gwynnda", month == "June", year == 2022) |> 
  pull(cigarette_butts) |> 
  sum(na.rm = TRUE)
```

For available data, Professor Trash Wheel collected a total of `r prof_total_weight` tons of trash. Gwynnda collected `r format(gwynnda_butts_june22, big.mark = ",")` cigarette butts in June of 2022.

# Problem 3

### 3.1 Data import + clean + questions
```{r}
baseline_df = 
  read_csv("data/MCI_baseline.csv", skip = 1, na = c("NA", "", ".")) |> 
  janitor::clean_names() |> 
  mutate(
    sex = recode_values(sex, 1 ~ "male", 0 ~ "female"),
    apoe4 = recode_values(apoe4, 1 ~ "carrier", 0 ~ "non_carrier")) |>
  filter(is.na(age_at_onset) | age_at_onset > current_age)

baseline_df

n_recruited = 
  read_csv("data/MCI_baseline.csv", skip = 1, na = c("NA", "", ".")) |> 
  nrow()

n_developed_mci = 
  baseline_df |> 
  filter(!is.na(age_at_onset)) |> 
  nrow()

prop_women_carriers = 
  baseline_df |> 
  filter(sex == "female") |> 
  mutate(carrier = apoe4 == "carrier") |> 
  pull(carrier) |> 
  mean()
```

Some of the crucial steps to import and clean this data was to indicate the rows and columns to be included (skipping the frist row that did not provided data entry) and specify that missing values should have NA as their entry. The variables sex and apoe4 needed to be recoded to avoid confusion with the meaning of 1 and 0. For this, two new variables where created (sex and apoe4) with the correct coding. Finally, the data was filtered to remove entries that did not meet the inclusion criteria. In other words, patients that had their MCI onset before the study began (age at onset < current age). 

In total, `r n_recruited` participants where recruited, of which `r n_developed_mci` developed MCI during follow-up. Among women in the study, `r round(prop_women_carriers * 100, 2)`% are APOE4 carriers.

### 3.2 Data import and clean of amyloid data set

```{r}

amyloid_df = 
  read_csv("data/mci_amyloid.csv", skip = 1, na = c("NA", "", ".", "Na")) |> 
  janitor::clean_names() |> 
  rename(id = study_id) |> 
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_ratio")

amyloid_df
```

This dataset contains the amyloid 42/40 ratio measured for each participant at up to five visits: baseline and 2, 3, 6, and 8 years after baseline. A few steps were needed to clean the data. The first row of the dile is a description and not the variables names, so it was skipped during import. Several observations had "NA" or "Na" entered in places of a ratio, and these were treated as missing. The variable `Study_id` was renamed to `id` to match the baseline dataset. The data were not tidy, since there weas a separate column for each visit, so I reorganized them to a long format with one row per participant per visit. The resulting dataset has `r nrow(amyloid_df)` rows and `r ncol(amyloid_df)` variables (`id`,`visit`, and `amyloid_ratio`), covering `r amyloid_df |> distinct(id) |> nrow()` participants. 

### 3.3 Checking whether participants appear in only one dataset
```{r}
baseline_only_df = 
  baseline_df |> 
  filter(!(id %in% pull(amyloid_df, id)))

amyloid_only_df = 
  amyloid_df |> 
  filter(!(id %in% pull(baseline_df, id))) |> 
  distinct(id)

nrow(baseline_only_df)
nrow(amyloid_only_df)
```

There are `r nrow(baseline_only_df)` participants that appear only on the baseline dataset and `r nrow(amyloid_only_df)` that appear only in the amyloid dataset. 

### 3.4 Combining dataset

```{r}
mci_df = inner_join(baseline_df, amyloid_df, by = "id")

write_csv(mci_df, "data/mci_combined.csv")
```

The combined dataset includes only the `r mci_df |> distinct(id) |> nrow()` participants who appear in both the baseline and amyloid datasets. It has `r nrow(mci_df)` rows and `r ncol(mci_df)` variables, with five rows per participant: one for each visit (baseline and years 2, 4, 6, and 8), identified by the variable `visit`. Each row gives the amyloid ratio measured at that visit (`amyloid_ratio`), along with the participant's baseline information (`current_age`, `sex`, `education`, `apoe4`, and `age_at_onset`), which repeats across their rows.


