---
title: "Homework 2"
author: "Andrea Vasquez"
date: " 10/04/2026"
output: 
  github_document:
    toc: true
---

# Problem 1

### 1.1 Data import and cleaning + brief description 

```{r}

library(tidyverse)

NYC_transit_df = 
  read_csv("./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
  na = c("NA", "", "."),
    col_types = cols(
      Route8 = col_character(),
      Route9 = col_character(),
      Route10 = col_character(),
      Route11 = col_character())) |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude,
         starts_with("route"), entry, vending, entrance_type, ada) |> 
  mutate(entry = case_match(entry, "YES" ~ TRUE, "NO" ~ FALSE))

NYC_transit_df

skimr::skim(NYC_transit_df)
```

The NYC transit dataset contains information on each subway entrance and exit. After cleaning, it has `r nrow(NYC_transit_df)`, rows and `r ncol(NYC_transit_df)` columns, of which 15 are character variables (line, station name, routes 1–11, vending, and entrance type), 2 are logical variables (entry and ADA compliance), and 2 are numeric variables (station latitude and longitude). To clean the datam, I imported the file specifying that blank cells be treated as missing and that routes 8-11 be read as character variables. These columns contain numbered trans and would otherwise be read as numeric, even though the values are rout name and not true numbers. I then cleaned the variables names, retained only the variables of interest, and converted the entry variable from character to logical. Given that route spread across 11 columns, we can conclude that this data is not tidy. To be tidy, there should be a single column for the variable route. 

### 1.2 Answering questions

#### Q1

```{r}
n_stations = 
  NYC_transit_df |>
  distinct(line, station_name) |>
  nrow()
```

There are `r n_stations` distinct stations. 

#### Q2

```{r}
n_ada_compliant =
  NYC_transit_df |>
  filter (ada == TRUE)|>
  distinct (line, station_name) |>
  nrow()
```

In total, `r n_ada_compliant` stations are complaint with ADA. 

#### Q3

```{r}
prop_entry_ven = 
  NYC_transit_df |>
  filter(vending == "NO") |>
  pull (entry) |>
  mean()
```

Among entrances/exits without vending, `r round(prop_entry_ven * 100, 2)`% allow entrance. 

### 1.3 Reformat data



